---
title: "Paper1"
author: "Group13"
date: "7 listopada 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(corrplot)
library(ResourceSelection)
```
## Part III (20 points)
Conduct a regression analysis and report your findings. The report should 
contain the following
sections:
• data import
• descriptive statistics and data validation
• identification of relevant regressors
• fitting a regression model
• discussion of model fit, e.g. goodness of fit, significance of regressors
• interpreting the model

#Data import and general information

```{r}
load(file = "titanic.Rdata")
#checking first row of the data frame
head(titanic)
#Size of dataframe
dim(titanic)
#structure of the data
str(titanic)
```
Our data set consists of 1309 rows=observations with 14 columns = feature variables. Observation is the particular passenger.
Now we want to get some summary statistics about each of variable
```{r}
#summary statistics for each variable
summary(titanic)
```
We check what data type each of variable is, for that purpose we use both command: class and typeof to compare the results
```{r}
sapply(titanic,typeof)
sapply(titanic, class)
```
We can see that our variables are correctly defined: pclass, sex are factors
#Data preparation
As we noticed many of the data has missing values, therefore we replace them with
NA. Furthermore 2 variables Name and Cabin. Name is a string which has no meaning 
for statistical purposes and cabin has too many empty entries (1019)

```{r}
#cleaning the data, impute NA for missing values to get consistency 
#of the dataset
data<-titanic
data[data==""]<-NA #insert NA for missing values
data$name<-NULL#deleting name from data frame
data$cabin<-NULL
#plotting for better uderstanding
datafactor<-data[1:3] #extract the factor/classification variables
plottingbar<- function (x){#function for plotting barplot
  barplot(prop.table(table(x)))
}
sapply(datafactor,plottingbar)
datanum<-data[4:7]#extract numeric variables
datanum$ticket<-NULL
sapply(datanum, hist)
#plot the connection
plot(data$age, data$survived)
plot(data$pclass, data$survived)
```
From the plot we may say that the male passengers were almost twice as more than female aboard, vast part of the passengers died and most of the passengers were 3rd class passengers.

#Descriptive analysis
Coorrelation matrix
```{r}
datacor<-data[1:8] #select the variables:pclass,survived,sex,age,sibsp,parch the rest seems to be irrelevant and hard to interpret
datacor$ticket<-NULL
datacor$cabin=NULL #delete cabin as it has too many na's
#chnaging pclass into integer
datacor$pclass<-as.character(datacor$pclass)
datacor$pclass[datacor$pclass=="1st"]<-1
datacor$pclass[datacor$pclass=="2nd"]<-2
datacor$pclass[datacor$pclass=="3rd"]<-3
datacor$pclass<-as.integer(datacor$pclass)
#changing sex into integer:1 for male, 0 for female
datacor$sex<-as.character(datacor$sex)
datacor$sex[datacor$sex=="male"]<-1
datacor$sex[datacor$sex=="female"]<-0
datacor$sex<-as.integer(datacor$sex)
#calculate correlation 
correlation<-cor(na.omit(datacor))
#plot the correlation
corrplot(correlation)
```
Summary of the correlations:
1. class and the survived is negatively correlated which means, the lower class the passenger was the less chance of survival one had
2.As we may expected before fare is negatively correlated with the pclass, so it means the the lower the price of the ticket the worse/lower class, which also proofs that the higher the fare it is more likely to survive due to positive relation between fare and survived
3.Taking look at the sex and survived we can notice that it is correlated negatively, in other words male passenger were more likely to die (having in mind that 1=male, 0=female)
# Some statistics about the data
```{r}
datatable<-na.omit(datacor[1:3])
table(datatable)
prop.table(table(datatable$pclass))
#write what the nr means for man and woman
paste('Number of passengers who survived the Titanics diseaster is', sum(data$survived))
```
Conclusions:
1.Most of the passengers (75-79%) from 1st and 2nd class survived, for third class only the half
2.Female passengers were more likely to survive than male
3.Most male passengers
  

```{r,echo=TRUE}
prop.table(table(datatable$survived))
```
38% of the passengers survived the titanic's diseaster
```{r}
prop.table(table(datatable$sex))
```
64% of the all passengers were male
```{r}
#function for checking how many na's are in a vector
nasearcher<-function(x){
  length(x[x=='NA'])
}
sapply(data, nasearcher)
#based on the function we decided that variables : ticket, boat, body and home destination are not used in the model due to too many missing values and not relevant
dataglm<-na.omit(subset(data, select=c(1,2,3,4,5,6,8,9)))
model<-glm(survived ~., family = binomial(link="logit"),data=dataglm)
summary(model)
```
**Interpretation:**
Variables Pclass2, Pclass3, sexmale and age are very statistically significant.
Only Parch and Fare show no signifcance.
Passenger Classes confirm the negative relation, the lower the class from 1st to
3rd the less likely is to survive. The same with sex, male are more likely to 
die and the older the passenger was the lower chances one had. For more precise interpretation the model coefficient should be transformed in exponential
#Recalculate the model
```{r}
#recalculate the model without parch and fare 
#as they are not statistically significant
dataglm2<-subset(dataglm, select=c(1,2,3,4,5,8))
model2<-glm(survived ~., family = binomial(link="logit"),data=dataglm2)
summary(model2)
#changing to exponential for better interpretation
exp(coef(model2))
```
**Final Interpretation**
Being a woman from 1st class with Embarked in Chambourg the odds of survival 
increases by a factor 74,8 (interpretation for intercept). Increase in age in 1 
year will decrease chance of survival by 0.96.
**Goodness of fit**
Looking at the deviance we can see that our model represents relatively good fit the residul deviance has reduced by 954.88 with loss of 7 degrees of freedom.No change with respect to the prevoius model
**AIC-Akaike Information Criterion**
The value of AIC has dropped sligthly from 74 to 70 which means that this model is better than previous one.
**Fisher Scoring**
It is an algorithm derived on Newton's method for solving maximum likelihood problems numerically,in our model 5 iterations were performed to reach the convergence so no change has been registered compared to the first model
#Plot of the model
```{r}
exp(cbind(OR = coef(model2), confint(model2)))
plot(model2)
```
**Hosmer-Lemeshow Goodness of Fit**
```{r}
hoslem.test(dataglm2$survived, fitted(model2))

```

